---
title: "Course Project 1"
author: "Andrew Chirwa"
date: "May 31, 2016"
output: html_document
---
###1. Read in the dataset and processing the data
```{r}
#Load Data
#setwd("C:/Coursera/reproducibleresearch/data")
activity <- read.csv("activity.csv", na.strings = "NA")

#Process/Transform Data
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity$interval <- factor(activity$interval)
```
###2. Histogram of the total number of steps taken each day
```{r}
#Ignore missing values
activity_without_NA <- activity[!is.na(as.character(activity$steps)),]

#Calculate total number of steps taken each day
sumofsteps<-tapply(activity_without_NA$steps, activity_without_NA$date, sum, na.rm=T)

#Make histogram of the total number of steps taken each day
hist(sumofsteps,breaks = 20, xlab="Number of steps", main = "Histogram of the total number of steps taken each day", col="red")  
```

###3. Mean and median number of steps taken each day  
```{r}
mean(sumofsteps)
median(sumofsteps)
```

###4. Time series plot of the average number of steps taken
```{r}
#Calculate the average
activity_interval <- aggregate(activity_without_NA$steps, by=list(interval=activity_without_NA$interval), FUN=mean)

#Add columns names
colnames(activity_interval) <- c("interval", "average_steps")

#Plot the average daily activity pattern 
plot(as.integer(levels(activity_interval$interval)), activity_interval$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="red")

```

###5. The 5-minute interval that, on average, contains the maximum number of steps
```{r}
max_num_steps <- max(activity_interval$average_steps)
max_num_steps

#interval_max_steps<-activity_interval[which.max(activity_interval$average_steps),]$interval
#interval_max_steps
activity_interval[which.max(activity_interval$average_steps),]$interval

```

##6. Strategy for imputing missing data
Missing values 
```{r}

sum(is.na(activity$steps))
```
Use the transform function to impute the missing data. 
  If steps is NA replace with the mean
  else replace with original value
Save the imputed data im imputed_data
```{r}

imputed_data <- transform(activity, steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
```

##7. Plot histogram for the total number of steps taken each day after missing values are imputed
```{r}
#Create a data frame with the steps taken for each day and add olumn names to the created data frame
steps_each_day_complete <- aggregate(steps ~ date, data = imputed_data, sum)
colnames(steps_each_day_complete) <- c("date", "steps")

#Make the histogram
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
mean(steps_each_day_complete$steps)
median(steps_each_day_complete$steps)
```

##8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
#Create a factor variable "day "to store the day of the week:
imputed_data$day <- as.factor(weekdays(imputed_data$date))

#Create a logical variable "is_weekday" (if weekday then TRUE, if weekend thenFALSE) :
imputed_data$is_weekday <- ifelse(!(imputed_data$day %in% c("Saturday","Sunday")), TRUE, FALSE) 

#Calculate the average number of steps for weekdays
weekdays_data <- imputed_data[imputed_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)

#Calculate the average number of steps for weekends
weekends_data <- imputed_data[!imputed_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)

#Add columns names
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")

#Add a column to indecate the day
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"

#Merge the two together
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)

#Convert the day variable to a factor -- "Weekday", "Weekend"
week_data$day <- as.factor(week_data$day)
```

Load the lattice package and plot the 2 graphs for the data
```{r}
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
```